High Cardinality Variable
===

### What is this about?

A **high cardinality** variable is one that can take _many_ different values. For example country. 

This chapter will cover some implications of not treating, and possible methods to deal with them.

```{r ,results="hide", echo=FALSE}
library(knitr)
opts_knit$set(base.dir = "data_preparation")
```
 
 
<br> 

### High Cardinality in Descriptive Statistics

The following example contains a survey of 1000 cases, with 2 columns: `country` and `flu`, which indicates having such illness in the last month.

```{r , message=F}
library(funModeling) 
```

`data_country` data comes inside `funModeling` package (please update to release 1.6).

Quick `data_country` profiling (first 10 rows)

```{r fig.height=6, fig.width=4}
# plotting first 10 rows
head(data_country, 10)

# exploring data, displaying only first 10 rows
head(freq(data_country, "country"), 10)
```

```{r fig.height=2, fig.width=2}
# exploring data
freq(data_country, "has_flu")
```

The last table shows there are **70 different countries**, and ~9% of people who had flu -`has_flu="yes"`.

But many of them have almost no participation in the data. This is the _long tail_, so one technique to reduce cardinality is to keep those categories that are present the a high percentahge of data share, for example 70, 80 or 90%, the Pareto principle.

```{r fig.height=3}
# 'freq' function, from 'funModeling' package, retrieves the cumulative_percentage that will help to do the cut. 
country_freq=freq(data_country, 'country', plot = F)

# Since 'country_freq' is an ordered table by frequency, let's inspect the first 10 rows with the most share.
country_freq[1:10,]
```


So 10 countries represent more the 70% of cases. We can assign the category `other` to the remaining cases and plot:

```{r}
data_country$country_2=ifelse(data_country$country %in% country_freq[1:10,'country'], data_country$country, 'other')
freq(data_country, 'country_2')
```

<br> 

### Final comments

Low representative categories are sometimes errors in data, such as having: `Egypt`, `Eggypt.`, and may give some evidence in bad habbits collecting data and/or possible errors when collecting from the source.

There is no general rule, just to shrink data, it depends on each case.

<br>

**Next recommended chapter: <a href="http://livebook.datascienceheroes.com/data_preparation/high_cardinality_descriptive_stats.html">High Cardinality Variable in Predictive Modeling</a>**




