Profiling Data
====


### What is this about?

Quantity of zeros, NA, Inf, unique values; as well as the data type may lead to a good or bad model. Here's an approach to cover the very first step in data modeling. 

```{r ,results="hide", echo=FALSE}
library(knitr)
opts_knit$set(base.dir = "data_preparation")
```

```{r lib, results="hide"}
## Loading funModeling !
suppressMessages(library(funModeling))
data(heart_disease)
```

### Checking NA, zeros, data type and unique values
```{r df_status}
my_data_status=df_status(heart_disease)
```
* `q_zeros`: quantity of zeros (`p_zeros`: in percentage)
* `q_inf`:  quantity of infinite values (`p_inf`: in percentage)
* `q_na`:  quantity of NA (`p_na`: in percentage)
* `type`: factor or numeric
* `unique`: quantity of unique values

#### Why are these metrics important?
* **Zeros**: Variables with **lots of zeros** may be not useful for modeling, and in some cases it may dramatically bias the model.
* **NA**: Several models automatically exclude rows with NA (**random forest**, for example). As a result, the final model can be biased due to several missing rows because of only one variable. For example, if the data contains only one out of 100 variables with 90% of NAs, the model will be training with only 10% of original rows.
* **Inf**: Infinite values may lead to an unexpected behavior in some functions in R.
* **Type**: Some variables are encoded as numbers, but they are codes or categories, and the models **don't handle them** in the same way.
* **Unique**: Factor/categorical variables with a high number of different values (~30), tend to do overfitting if categories have low cardinality, (**decision trees**, for example).

#### Filtering unwanted cases
The function `df_status` takes a data frame and returns a the status table to quickly remove unwanted cases.


**Removing variables with high number of NA/zeros**
```{r df_status3}
# Removing variables with 60% of zero values
vars_to_remove=subset(my_data_status, my_data_status$p_zeros > 60)
vars_to_remove["variable"]

## Keeping all except vars_to_remove
heart_disease_2=heart_disease[, !(names(heart_disease) %in% vars_to_remove[,"variable"])]

```

**Ordering data by percentage of zeros**
```{r df_status4}
my_data_status[order(-my_data_status$p_zeros), c('variable', 'p_zeros')]
```

<br>


### Profiling character/factor variable

Frequency or distribution analysis made it simple by `freq` function. It retrieves the distribution in a table and a plot (by default) which shows the distribution in absolut and relative numbers.

If you want the distribution for two variables: 

```{r fig.height=3, fig.width=5}
freq(data=heart_disease, str_input = c('thal','chest_pain'))
```

As well as remaining `funModeling` functions, if `str_input` is missing it will run for all factor or character variables present in given data frame:

```{r, eval=F}
freq(data=heart_disease)
```
<br>

Also, as the other plot function in the package, there is the need of exporting plots, add the `path_out` parameter (it will create the folder if it's not created yet)

```{r, eval=F}
freq(data=heart_disease, path_out='my_folder')
```

