# Data Science Live Book

```{r ,results="hide", echo=FALSE}
library(knitr)
opts_knit$set(base.dir = "../readme")
```

A book to learn data science, data analysis and machine learning, suitable for all ages!

## Introduction

This book covers common aspects in predictive modeling:

+  A. **Data Preparation** / **Data Profiling**
+  B. **Selecting best variables (dataviz)**
+  C. **Assessing model performance**
+  D. **Miscellaneous**

And it is heavly based on the `funModeling` package. Please install before starting:

`install.packages("funModeling")`


As you can see, model creation is not included, while _almost_ everything else does. 

Model creation consumes around **10%** of almost any predictive modeling project; `funModeling` will try to cover remaining 90%. 

It´s not only the function itself, but the explanation of how to interpret results. This brings a deeper understanding of **what is being done**, boosting the freedom to use that knowledge in other situations andor using languages.


## `funModeling` quick start

```{r, results="hide"}
## Loading funModeling !
suppressMessages(library(funModeling))

# loading data frame
data(heart_disease)
```

### Part A: Data Preparation / Profiling

When a new data set arrives...:

```{r}
df_status(heart_disease)
```

And if there are some outliers... top or bottom values can be removed using:
```{r, results="hide"}
data_2=prep_outliers(data = heart_disease,  type='set_na', top_percent  = 0.01)
```



##### Full info at: 





### Part B: Selecting best variables (dataviz)

**Overview:** `cross_plot` discretizes input variable automatically exposing thei relationship against a target variable.

```{r,  fig.height=3, fig.width=7}
cross_plot(data=heart_disease, str_input="age", str_target="has_heart_disease")
```

```{r,  fig.height=3, fig.width=5}
plotar(data=heart_disease, str_input="age", str_target="has_heart_disease", plot_type="histdens")

```

`plotar` and `cross_plot` can plot all numerical variables at once, as well as export each graph to a high quality _png image_.

##### Full info at: 

### Part C: Assessing model performance

Full info at: 

### Part D: Miscellaneous

##### Full info at: 


